# (PART\*) Appendix {-} 

# Assignments {-} 

## Presentation {-} 

The lecture and the associated literate cover a range of topics. Those topics will be assigned to one student or a small group of students. Those assignees are responsible for the topic and prepare an educational presentation that 1) helps their fellow students to quickly recapitulate the basic concepts associated with the topic 2) highlights and explains potential pitfalls that are expected to appear when applying the concepts of the topic and 3) introduce at least one example that shows the application of the respective context in an analytical environment. The whole has to be in the form of a **presentation**, but especially the application can be supported with code either directly embedded in the presentation or in the form of an additional code file or notebook.

## Dataset {-} 

Identifying a data set is a preliminary step to prepare the upcoming seminar paper. Every seminar paper needs an **individual** data set. The data set can cover any topic as long as it supports the application of the methodological approaches discussed in the course. This effectively requires the data set to exhibit a metric dependent variable that can be explained by a set of independent variables (features) according to some formulated theory. The source of the data set can be any eligible source that supports academic usage, such as a company that you are associated with, scraped from a website or a collection of websites, an existing research paper that published its data or a data archive. There are no real constraints regarding the number of variables or number of observations as long as the data set is large enough to convincingly answer a formulated research hypothesis. Some examples for data sets can be found in the `fhswf` R package:

```{r, eval=FALSE}
library(fhswf)
?fhswf::boatprices # Explanation of boat prices using features of the boat.
?fhswf::whodata    # Explanation of macroeconomic indicators such as the 
                   # human development index of a country using descriptives
                   # for that country incl. considering development over time.
```

## Seminar Paper {-} 

The seminar paper presents a comprehensive analytical use case that uses the proposed (and approved) data set to apply all (necessary) methodological steps teached and discussed in the course and the respective literature. This seminar paper is a piece of academic literature and therefore at least one research hypothesis needs to be formulated. This research hypothesis needs to be derived from literature (especially academic papers) and supporting arguments in form a formulated theory need to be presented. Next the data set needs to be introduced including univariate descriptive statistics for all relevant variables as well as statistics that explore the association of the introduced variables. The core of the seminar paper is the actual data analysis that is used to provide evidence for the stated research hypothesis. This analysis is followed by a intensive interpretation that deals with the topic of the paper (not statistical methodology) before you conclude, summarize and critique your work. This seminar paper deals with some form of (real world) domain, so the main arguments should relate to the chosen topic. As a consequence the employed statistical methodology requires no additional explanation. It can be assumed that all methodological topics that are actively teached in the course are well understood by the reader of your paper. However, if you choose to employ additional methods, more comprehensive statistical tests or any form of more complex modeling techniques, they need to be (briefly) introduced in the form of an additional *methodological theory* section to foster a common understanding. In summary the seminar paper is used to *mine* new knowledge from the presented data for an audience that is mainly interested in those findings and the domain of the data. This means the main body of the seminar paper consists of text and data, models or results are preferably presented in tables opposed to console output. As a consequence the actual code to generate the results is located in the appendix of the paper and not part of the body of the paper.

# Examples {-} 

Regression analysis is one of the most popular tools for academic research and is used frequently in almost all fields of science. Depending on the topic and the details of a specific research project the corresponding articles quickly become complex and difficult to understand. This is especially true, when the matter they discuss is complex and intertwined on its own. 

For the sake of simplicity the following example papers discuss a rather intuitive concept: Apples. The papers referenced below investigate and analyse multiple perspectives on the prices and quality characteristics of apples. They differ by age, depth of methodological expertise, research community, cultural background and of course quality. Therefore these examples are no strict guideline or template but should provide a sneak peek of how regression results are derived, discussed and displayed in academic practice. 

- Blend, J. R., & van Ravenswaay, E. O. (1999). Measuring Consumer Demand for Ecolabeled Apples. American Journal of Agricultural Economics, 81(5): 1072–1077. https://doi.org/10.2307/1244086
- Carew, R. (2000). A Hedonic Analysis of Apple Prices and Product Quality Characteristics in British Columbia. Canadian Journal of Agricultural Economics/Revue canadienne d'agroeconomie, 48: 241-257. https://doi.org/10.1111/j.1744-7976.2000.tb00278.x
- Carew, R. and Smith, E.G. (2004). The value of apple characteristics to wholesalers in western Canada: A hedonic approach. Canadian Journal of Plant Science. 84(3): 829-835. https://doi.org/10.4141/P03-164
- Kajikawa, C. (1998). Quality level and price in Japanese apple market. Agribusiness, 14: 227-234. [https://doi.org/10.1002/(SICI)1520-6297(199805/06)14:3<227::AID-AGR5>3.0.CO;2-2](https://doi.org/10.1002/(SICI)1520-6297(199805/06)14:3<227::AID-AGR5>3.0.CO;2-2)
- Khan, J., S.A. Shah, K.N. Saddozai, M. Fayaz, S. Ali, A.U. Jan and G. Ali. (2019). A hedonic price analysis of consumers preferences and willingness to pay for quality attributes of apple. Sarhad Journal of Agriculture, 35(4): 1243-1246. http://dx.doi.org/10.17582/journal.sja/2019/35.4.1243.1246
- Seppä, L., Latvala, T., Akaichi, F., Gil, J. M., & Tuorila, H. (2015). What are domestic apples worth? Hedonic responses and sensory information as drivers of willingness to pay. Food quality and preference, 43: 97-105. https://doi.org/10.1016/j.foodqual.2015.02.013

Depending on the academic outlet and the research community the length and depth of the methological discussions varies greatly. However, each of these papers relies on some statistical software package and a body of statistical tests that is larger than whats presented in the final publication. This - potentially not published - part of the work can be presented in an additional appendix made available solely for the reviewers. When writing a seminar paper those methodological additions should of course be part of the submission. 

# System Info {-}

```{r}
Sys.time()
sessionInfo()
```

